---
title: "PelageGenes"
author: "Enrico"
date: "28 May 2018"
output: html_document
---

# 0. Path Definition

Define the paths to different locations for easier calling during script

```{r Define paths, eval=FALSE, engine='bash'}

ssh -X ebazzicalupo@genomics-b.ebd.csic.es # these analyses will be conducted in server B because the annotated VCF file is located there

Proj_PATH=/home/ebazzicalupo/PelageGenes # path to the project directory
Data_PATH=/home/ebazzicalupo/Data # path to data directory
LPRef_PATH=/home/GRUPOS/grupolince/reference_genomes/lynx_pardinus_genome/lp23_plus_mt.fa # path to lynx pardinus reference genome file
VCF_PATH=/home/GRUPOS/grupolince/lynx_genomes_5x/VCFs_Dani/c_ll_lp_plus_h_ll_aafilled_SNPs_standard_filter.ann.vcf # path to annotated VCF file
BDB_Path=/home/ebazzicalupo/Data/BLASTdatabase # path to BLAST database
DiscBDB_PATH=/home/ebazzicalupo/Data/Discarded_BLASTdatabase # path to Discarded Scaffold BLAST database

```


# 1. Search for Genes in Lynx pardinus Genome Annotations

Look for the genes names/abbreviations in the genome annotations files. First look in the Lynx pardinus genome annotations. If any are not found, look for them in Felis catus genome annotations [step 2], then in Homo sapiens genome annotations [step 3].

```{r Search in Lynx Annotations, eval=FALSE, engine='bash'}

## A file called PelageGeneList.txt has been prepared with the gene names written for a boolean search through the grep command. Each line is a gene. The file has been located in the project directory (/home/ebazzicalupo/PelageGenes).

## Genome annotation files for the three species have also been copied in the Data directory (/home/ebazzicalupo/Data). The Felis catus and Homo sapiens Genome Annotations have been downloaded from the ensembl website (https://www.ensembl.org/info/data/ftp/index.html).

cd $Proj_PATH

mkdir LynxResults # create directory to store the results for Lynx

GENES=$(cat PelageGeneList.txt) # call all elements of PelageGeneList.txt

IFS=$'\n' # define the end of a line as a separator for elements so that each ENTIRE line in PelageGeneList.txt will be a search entry

# for each gene create a file containing results and move it to results directory
for i in ${GENES[@]}
  do
  grep -i -E "${i}" $Data_PATH/LYPA23C.all.fix.gff3 > ${i}
  mv ${i} LynxResults
done


```

Take the results and find the Parent Gene ID for each. Create a .txt file with the list of all Parent Gene IDs found.

```{r Refine Lynx Results, eval=FALSE, engine='bash'}

cd $Proj_PATH/LynxResults

mkdir ParentIDs # create a directory for storing the Parent Gene ID file

LYNXREFINE=$(ls *) # list all results

# for each result, cut the parent gene ID, sort it, eliminate duplicates and create a .Parent file with the Parent Gene IDs stored inside
for i in ${LYNXREFINE[@]}
  do cut -d ';' -f2 "${i}" | sort -u > "${i}".Parent
  mv "${i}".Parent ParentIDs
done

cd $Proj_PATH/LynxResults/ParentIDs

cat * > ParentIDs.txt # put Parent Gene IDs from all the results in a single .txt file

mv ParentIDs.txt $Proj_PATH

```

# 2. Search in Felis cauts Genome Annotations for Genes not found in Lynx pardinus Genome Annotations. Extract corresponding sequence FASTA. Run BLAST to Lynx pardinus Reference Genome.

Repeat the search in Felis catus genome annotations, using as input the empty results from the Lynx search [step 2a]. This will give coordinates to specific regions in the Felis catus genome.
Use those coordinates to extract the corresponding sequence (.fa) from the Felis catus reference genome [step 2b], and BLAST it to the Lynx pardinus reference genome [step 2c]. Hit results from the BLAST will be analysed individually [step 4], to see if any genes in Lynx pardinus can be added to the Parent Gene ID list for further analyses [steps 5].

## 2a - Search in Cat annotations

```{r Search in Cat Annotations, eval=FALSE, engine='bash'}

cd $Proj_PATH

mkdir CatResults # create directory to store the results for Felis catus

CATGENES=$(wc -l LynxResults/* | grep ' 0 ' | cut -d '0' -f2 | sed 's/ //' | sed 's/LynxResults\///') # List empty results from Lynx search

IFS=$'\n' # define the end of a line as a separator for elements so that each ENTIRE line in PelageGeneList.txt will be a search entry

# for each gene create a file containing results and move it to results directory
for i in $CATGENES; do grep -i -E "${i}" $Data_PATH/Felis_catus.Felis_catus_8.0.92.gff3 > cat_"${i}".gff3; mv cat_"${i}".gff3 CatResults; done

```

## 2b - Generate Cat gene FASTA files for BLAST

```{r Generate Cat gene fasta files for BLAST, eval=FALSE, engine='bash'}

## Reference Genome FASTA files for Felis catus and Homo sapiens have been downloaded from the ensembl website (https://www.ensembl.org/info/data/ftp/index.html) and copied in the Data directory (/home/ebazzicalupo/Data).

cd $Proj_PATH/CatResults

CATREFINE=$(ls cat_*.gff3) # list results from grep search

# for each result from the grep search, create a gene.gff3 file with the non-gene (e.g. mRNA, CDS) strings removed
for i in ${CATREFINE[@]}
  do
  awk -v "key=gene" '$3 == key {print($0)}' ${i} > ${i/.gff3}.gene.gff3
done


CATcoords=$(ls cat_*.gene.gff3) # list gene.gff3 files

# generate FASTA file for each gene.gff3 coordinates file
for i in ${CATcoords[@]}
  do
  bedtools getfasta -fo "${i/.gene.gff3}".fa -fi $Data_PATH/Felis_catus.Felis_catus_8.0.dna_sm.toplevel.fa -bed ${i}
done

```

## 2c - BLAST cat genes to Lynx genome

```{r  BLAST cat genes to Lynx genome, eval=FALSE, engine='bash'}

## First create a BLASTdatabase for the Lynx pardinus FASTA file

cd $Data_PATH

mkdir BLASTdatabase # create a directory for the BLAST database

BDB_Path=/home/ebazzicalupo/Data/BLASTdatabase # define path to BLAST database

scp $LPRef_PATH $BDB_Path # copy reference genome to the BLASTdatabase directory

# create a BLAST database in the BLASTdatabase directory
makeblastdb -in $BDB/lp23_plus_mt.fa -parse_seqids -dbtype nucl

## BLAST each Cat Gene FASTA to the Lynx pardinus Reference Genome and BLAST database. Look at BLAST manual for output options.

cd $Proj_PATH/CatResults

IFS=$'\n'

CBLASTGENES=$(ls cat_*.fa) # list Cat Gene FASTA files

for GENE in ${CBLASTGENES[@]}
  do
  echo "$GENE starting BLAST"
  blastn -query $Proj_PATH/CatResults/${GENE} -db $BDB_Path/lp23_plus_mt.fa -outfmt "6 qseqid sseqid sstart send length qcovhsp evalue" -out ${GENE/.fa/}.BLASTresults
  echo "$GENE BLAST finished"
done

## Move the BLASTresults to a new directory

mkdir BLASTresults

mv *.BLASTresults BLASTresults

```

## 2d - BLAST cat genes to Lynx discarded scaffolds

```{r  BLAST cat genes to Lynx discarded scaffolds, eval=FALSE, engine='bash'}

## Create BLAST database for discarded scaffold

cd $Data_PATH

mkdir Discarded_BLASTdatabase # create a directory for the Discarded Scaffold BLAST database

DiscBDB_PATH=/home/ebazzicalupo/Data/Discarded_BLASTdatabase # define path to Discarded Scaffold BLAST database

scp home/GRUPOS/grupolince/reference_genomes/lynx_pardinus_genome/lp23.discarded.scaffolds.fa $DiscBDB_PATH # copy Discarded Scaffold to the Discarded_BLASTdatabase directory

# create a BLAST database in the Discarded_BLASTdatabase directory
makeblastdb -in $DiscBDB_PATH/lp23.discarded.scaffolds.fa -parse_seqids -dbtype nucl


## BLAST each Cat Gene FASTA to the Lynx pardinus Discarded Scaffold BLAST database. Look at BLAST manual for output options.

cd $Proj_PATH/CatResults

CDSBLASTGENES=$(ls cat_*.fa) 

IFS=$'\n'

for GENE in ${CDSBLASTGENES[@]}
  do
  echo "$GENE starting BLAST"
  blastn -query $Proj_PATH/CatResults/${GENE} -db $DiscBDB_PATH/lp23.discarded.scaffolds.fa -outfmt "6 qseqid sseqid sstart send length qcovhsp evalue" -out ${GENE/.fa/}.DiscBLASTresults
  echo "$GENE BLAST finished"
  
done

## Move BLASTresults to new directory

mkdir DiscBLASTresults

mv *.DiscBLASTresults DiscBLASTresults

```

# 3. Search in Homo sapiens Genome Annotations for Genes not found in Lynx pardinus and Felis catus Genome Annotations. Extract corresponding sequence FASTA. Run BLAST to Lynx pardinus Reference Genome.

Repeat the search in Homo Sapiens genome annotations, using as input the empty results from the Felis catus search [step 3a]. This will give coordinates to specific regions in the Homo sapiens.
Use those coordinates to extract the corresponding sequence (.fa) from the Homo sapiens reference genome [step 3b], and BLAST it to the Lynx pardinus reference genome [step 3c]. Hit results from the BLAST will be analysed individually, to see if any genes in Lynx pardinus can be added to the Parent Gene ID list for further analyses [steps XX].

## 3a - Search in Human annotations

```{r Search in Human Annotations, eval=FALSE, engine='bash'}

cd $Proj_PATH

mkdir HumanResults # create directory to store the results for Homo sapiens

HUMANGENES=$(wc -l CatResults/*.gene.gff3 | grep ' 0 ' | cut -d '0' -f2 | sed 's/CatResults\/cat_//g' | sed 's/ //') # List empty results from Cat search

# for each gene create a gff3 file containing results and move it to HumanResults folder
for i in ${HUMANGENES[@]}
  do
  grep -i "${i/.gene.gff3/}" $Data_PATH/Homo_sapiens.GRCh38.92.gff3 > human_${i/.gene.gff3/}.gff3
  mv human_${i/.gene.gff3/}.gff3 HumanResults
done

```

## 3b - Generate Human gene FASTA files for BLAST

```{r Generate Human gene fasta files for BLAST, eval=FALSE, engine='bash'}

## Reference Genome FASTA files for Felis catus and Homo sapiens have been downloaded from the ensembl website (https://www.ensembl.org/info/data/ftp/index.html) and copied in the Data directory (/home/ebazzicalupo/Data).

cd $Proj_PATH/HumanResults

HUMANREFINE=$(ls human_*.gff3) # list results from grep search

# for each result from the grep search, create a gene.gff3 file with the non-gene (e.g. mRNA, CDS) strings removed
for i in ${HUMANREFINE[@]}
  do
  awk -v "key=gene" '$3 == key {print($0)}' ${i} > ${i/.gff3}.gene.gff3
done

HUMANcoords=$(ls human_*.gene.gff3) # list gene.gff3 files

# generate FASTA file for each gene.gff3 coordinates file
for i in ${HUMANcoords[@]}
  do
  bedtools getfasta -fo "${i/.gene.gff3}".fa -fi $Data_PATH/Homo_sapiens.GRCh38.dna_sm.toplevel.fa -bed ${i}
done

```

## 3c - BLAST human genes to Lynx genome

```{r  BLAST human genes to Lynx genome, eval=FALSE, engine='bash'}

## BLAST each Human Gene FASTA to the Lynx pardinus Reference Genome and BLAST database. Look at BLAST manual for output options.

cd $Proj_PATH/HumanResults

IFS=$'\n'

HBLASTGENES=$(ls human_*.fa) # list Human Gene FASTA files

for GENE in ${HBLASTGENES[@]}
  do
  echo "$GENE starting BLAST"
  blastn -query $Proj_PATH/HumanResults/${GENE} -db $BDB_Path/lp23_plus_mt.fa -outfmt "6 qseqid sseqid sstart send length qcovhsp evalue" -out ${GENE/.fa/}.BLASTresults
  echo "$GENE BLAST finished"
done

## Move the BLASTresults to a new directory

mkdir BLASTresults

mv *.BLASTresults BLASTresults

```

## 3d - BLAST human genes to Lynx discarded scaffolds

```{r  BLAST human genes to Lynx discarded scaffolds, eval=FALSE, engine='bash'}

## BLAST each Cat Gene FASTA to the Lynx pardinus Discarded Scaffold BLAST database. Look at BLAST manual for output options.

cd $Proj_PATH/HumanResults

HDSBLASTGENES=$(ls human_*.fa) 

IFS=$'\n'

for GENE in ${HDSBLASTGENES[@]}
  do
  echo "$GENE starting BLAST"
  blastn -query $Proj_PATH/HumanResults/${GENE} -db $DiscBDB_PATH/lp23.discarded.scaffolds.fa -outfmt "6 qseqid sseqid sstart send length qcovhsp evalue" -out ${GENE/.fa/}.DiscBLASTresults
  echo "$GENE BLAST finished"
  
done

## Move BLASTresults to new directory

mkdir DiscBLASTresults

mv *.DiscBLASTresults DiscBLASTresults

```

# 4. Copy BLASTresults files from server to the laptop for consultation. Additional genes discovered during this step should be integrated to the ParentIDs.txt file before step 5

```{r  copy BLASTresults files from server to laptop, eval=FALSE, engine='bash'}

# cat greps
scp ebazzicalupo@genomics-b.ebd.csic.es:/home/ebazzicalupo/PelageGenes/CatResults/*.gff3 ~/home/ebazzicalupo/GrepResults

# cat coordinates
scp ebazzicalupo@genomics-b.ebd.csic.es:/home/ebazzicalupo/PelageGenes/CatResults/BLASTresults/cat_*.BLASTresults ~/home/ebazzicalupo/Results

# cat discaffold coordinates
scp ebazzicalupo@genomics-b.ebd.csic.es:/home/ebazzicalupo/PelageGenes/CatResults/DiscBLASTresults/* ~/home/ebazzicalupo/Results

# human greps
scp ebazzicalupo@genomics-b.ebd.csic.es:/home/ebazzicalupo/PelageGenes/HumanResults/*.gff3 ~/home/ebazzicalupo/GrepResults

# human coordinates
scp ebazzicalupo@genomics-b.ebd.csic.es:/home/ebazzicalupo/PelageGenes/Data/HumanResults/BLASTresults/human_*.BLASTresults ~/home/ebazzicalupo/Results

# human discaffold coordinates
scp ebazzicalupo@genomics-b.ebd.csic.es:/home/ebazzicalupo/PelageGenes/HumanResults/DiscBLASTresults/* ~/home/ebazzicalupo/Results

```

# 5. Intersect the VCF data with the Gene coordinates from the previous search to create a VCF with the subset of SNPs that are found within the Gene coordinates regions

Using the ParentIDs.txt file, generate a BED file containing the coordinates of all the genes [step 5a]. Perform some editing in order to make the BED file compatible with the VCF file that will be used. A VCF to which detailed annotations have been previously added will be used. Intersect the VCF file with the BED to obtain a VCF format of the list of all SNPs falling in the regions specified by the BED, with detailed information regarding the region [step 5b].

## 5a - Create BED from ParentIDs and edit it to be compatible with the VCF

```{r  Create BED from ParentIDs and edit for compatibility, eval=FALSE, engine='bash'}

cd $Proj_PATH

# create GFF3 file by looking for ParentIDs in Lynx Genome Annotations
grep -f ParentIDs.txt $Data_PATH/LYPA23C.all.fix.gff3 > Candidate.gff3

# create BED file from the GFF3 with the following tab separated columns: scaffold, start, end, type, info
cut -f1,3,4,5,9 Candidate.gff3 | awk 'BEGIN {FS="\t"; OFS="\t"} {print $1, $3, $4, $2, $5}' > Candidate.bed

# modify BED to allow VCF compatibility (scaffold name matching) - sort and insert a . at 4th character in each line
sort -k1,2 Candidate.bed | awk -vFS="" -vOFS="" '{$4=$4"."}1' > Candidate.compatible.bed

```

## 5b - Intersect compatible BED with the annotated VCF and remove repetitions generated by the overlapping regions in the BED file

```{r  Intersect BED with the VCF and remove repetitions, eval=FALSE, engine='bash'}

## Intersect BED with the VCF

cd $Proj_PATH

mkdir VCF # create directory for VCF results

# intersect the VCF file with the compatible BED
bedtools intersect -header -a $VCF_PATH -b Candidate.compatible.bed > $Proj_PATH/VCF/lp_bedIntersect.vcf

## Remove duplicates from the Intersect file

cd $Proj_PATH/VCF

grep "#" lp_bedIntersect.vcf > lp_bedIntersect_unique.vcf # copy header part

grep -v "#" lp_bedIntersect.vcf | sort -u -k1,2 >> lp_bedIntersect_unique.vcf # copy table, remove duplicates, add it to header

```

